$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-04-29 paper/paper.tex
Give the operational semantics another pass

  \inferrule
  {uQ : \Unused_0 \, Q}
  {\constr{scope-ext} : \new (\comp{P}{Q}) \eqeq \comp{(\new P)}{lower \; 0 \; \; Q \; uQ}}

  \inferrule
  { }
  {\constr{scope-comm} : \new \new P \eqeq \new \new swap \; 0 \; P}
\end{mathpar}
\caption{Base cases of structural congruence.}
\label{fig:struct-cong-base}
\end{figure}

@@

Figure \ref{fig:reduction} models the operational semantics of the \picalc{}.
Processes put in parallel reduce when they communicate over a common variable.
The receiving process substitutes references to its most immediate variable with references to the variable sent by the process doing the output.
The process is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under structural congruence and goes under parallel composition and scope restriction
--- but notably not under input nor output, otherwise the sequencing of actions would not be preserved.
In \S \ref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.
However, this reduction operation is effectful: it consumes the variable over which communication happens.
If this variable is external to $P$ (it resides in its context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the variable index over which communication occurs to the type level.
Every time we come out of a binder we decrement this variable.
To do so, we make use of a $dec$ function that saturates at $\constr{internal}$ --- as opposed to an $inc$ function that would not.

@@

  \inferrule
  {i \; j : \Var_n \\ P : \Process_{\suc n} \\ Q : \Process_n}
  {\constr{comm} : \comp{\recv{i}P}{\send{i}{j}{Q}} \reduce{\constr{external} \; i} \comp{lower \; 0 \; (\subst{P}{j}{0})}{Q}}

@@

\subsection{Framing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Weakening}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Strengthening}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{nitheorem}
  \label{thm:strengthening}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index that is not mentioned by $P$, such that $\Unused \; i \; P$.
  Then removing index $i$ from the typing context and both of the usage contexts and decrementing every variable in $P$ greater than $i$ preserves its well typedness, so that $\types{delete_i \; \gamma}{delete_i \; \Gamma}{lower_i \; P}{delete_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\subsection{Swapping}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Substitution}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Subject Reduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{proof}
  \hfill{}\\
  \begin{itemize}
    \item
    In the base case $\constr{comm}$ we need to construct a typing derivation $\types{\gamma}{\Gamma}{lower \; 0 \allowbreak \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi}$ starting from the assumptions in Figure \ref{fig:subject-reduction:a}.
    We start by using framing (Theorem \ref{thm:framing}) to rearrange the consumption of the input and output multiplicities of the communicating channel and then apply substitution (Theorem \ref{thm:substitution}) and strengthening (Theorem \ref{thm:strengthening}).

    \begin{figure}[h]
      \setlength{\mathindent}{0pt}
      \begin{subfigure}{.3\textwidth}
        \begin{alignat*}{2}
          \containsusage{&\Gamma'}{i}{\ell_\# &&}{\Gamma} \\
          \contains{\gamma}{&\Gamma'}{i}{\channel{t}{m}}{\ell_i &&}{\Psi} \\
          \types{\gamma , t}{&\Psi , m}{P &&}{\Delta , \ell_{\o}} \\
          \contains{\gamma}{&\Delta }{i}{\channel{t}{m}}{\ell_o &&}{\Theta} \\
          \contains{\gamma}{&\Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Initial assumptions.}
        \label{fig:subject-reduction:a}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{alignat*}{2}
          \types{\gamma , t}{& \Gamma , m}{P &&}{\Theta , \ell_{\o}} \\
          \contains{\gamma}{& \Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Rearrange $\ell_o$ witness, and merge with $\ell_i$ witness into $\ell_\#$.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma , t}{\Gamma , m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi , m}
        \end{align*}
        \caption{Substitution.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma , t}{\Gamma , m}{lower \; 0 \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi , m}
        \end{align*}
        \caption{Strengthening.}
      \end{subfigure}
      \caption{Proof steps for the $\constr{comm}$ case in subject reduction.}
    \end{figure}

>>>>

  \inferrule
  {uQ : \Unused_0 \, Q}
  {\constr{scope-ext} : \new (\comp{P}{Q}) \eqeq \comp{(\new P)}{lower_0 \; \; Q \; uQ}}

  \inferrule
  { }
  {\constr{scope-comm} : \new \new P \eqeq \new \new swap_0 \; P}
\end{mathpar}
\caption{Base cases of structural congruence.}
\label{fig:struct-cong-base}
\end{figure}

@@

Figure \ref{fig:reduction} models the operational semantics of the \picalc{}.
Processes put in parallel reduce when they communicate over a common channel.
The receiving process substitutes references to its most immediate variable with references to the variable that is being sent by the process doing the output.
The process is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under structural congruence and goes under parallel composition and scope restriction
--- notably, not under input nor output, or otherwise the sequencing of actions would not be preserved.
In \S \ref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.
However, this reduction operation is effectful: it consumes the channel over which communication occurs.
If this channel is external to $P$ (it resides in its context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the index of the channel to the type level.
We use a $dec$ function to decrement the index of the channel every time we come out of a binder.
This function saturates at $\constr{internal}$ --- meaning communication happens internally to the process.

@@

  \inferrule
  {i \; j : \Var_n \\ P : \Process_{\suc n} \\ Q : \Process_n}
  {\constr{comm} : \comp{\recv{i}P}{\send{i}{j}{Q}} \reduce{\constr{external} \; i} \comp{lower_0 \; (\subst{P}{j}{0})}{Q}}

@@

\subsection{Framing}
\label{framing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Weakening}
\label{weakening}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Strengthening}
\label{strengthening}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{nitheorem}
  \label{thm:strengthening}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index that is not mentioned by $P$, such that $\Unused_i \; P$.
  Then removing index $i$ from the typing context and both of the usage contexts and decrementing every variable in $P$ greater than $i$ preserves its well typedness, so that $\types{delete_i \; \gamma}{delete_i \; \Gamma}{lower_i \; P}{delete_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\subsection{Swapping}
\label{swapping}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Substitution}
\label{substitution}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Subject Reduction}
\label{subject-reduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{proof}
  \hfill{}\\
  \begin{itemize}
    \item
    In the base case $\constr{comm}$ we need to construct a typing derivation $\types{\gamma}{\Gamma}{lower_0 \allowbreak \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi}$ starting from the assumptions in Figure \ref{fig:subject-reduction:a}.
    We start by using framing (Theorem \ref{thm:framing}) to rearrange the consumption of the input and output multiplicities of the communicating channel and then apply substitution (Theorem \ref{thm:substitution}) and strengthening (Theorem \ref{thm:strengthening}).

    \begin{figure}[h]
      \setlength{\mathindent}{0pt}
      \begin{subfigure}{.3\textwidth}
        \begin{alignat*}{2}
          \containsusage{&\Gamma'}{i}{\ell_\# &&}{\Gamma} \\
          \contains{\gamma}{&\Gamma'}{i}{\channel{t}{m}}{\ell_i &&}{\Psi} \\
          \types{\gamma , t}{&\Psi , m}{P &&}{\Delta , \ell_{\o}} \\
          \contains{\gamma}{&\Delta }{i}{\channel{t}{m}}{\ell_o &&}{\Theta} \\
          \contains{\gamma}{&\Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Initial assumptions.}
        \label{fig:subject-reduction:a}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{alignat*}{2}
          \types{\gamma , t}{& \Gamma , m}{P &&}{\Theta , \ell_{\o}} \\
          \contains{\gamma}{& \Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Rearrange $\ell_o$ witness, and merge with $\ell_i$ witness into $\ell_\#$.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma , t}{\Gamma , m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi , m}
        \end{align*}
        \caption{Substitution.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma , t}{\Gamma , m}{lower_0 \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi , m}
        \end{align*}
        \caption{Strengthening.}
      \end{subfigure}
      \caption{Proof steps for the $\constr{comm}$ case in subject reduction.}
    \end{figure}
